---
title: "Annotation cluster k20"
author: "NadineBestard"
date: "30/03/2021"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE, warning=FALSE}
library(here) #reproducible paths
library(scater) #feature plots
library(patchwork) # agregate plots
library(scran) # for findmarkers
library(pals) # for palettes with large n #kelly()22, #polychrome()#36, unname(alphabet())
```

```{r age}
age <- "old"
```
```{r pals}
# remove the black and white from the pallete, still 20 colours left
kelly_col <- unname(kelly()[-c(1,2)])
```

```{r import}
if (!file.exists(here("processed", age, "sce_anno_02.RDS"))) {
  sce <- readRDS(here("processed", age, "sce_clusters_02.RDS"))
}else{
  sce <- readRDS(here("processed", age, "sce_anno_02.RDS"))
}
```

## Annotation

We make a first rough annotation using known markers for different
celltypes.


```{r}
plotTSNE(sce, colour_by = "cluster_k20", text_by = "cluster_k20") + scale_color_manual(values = unname(alphabet()))
```

<details>

<summary> Click to expand the Neurons marker plots </summary>

```{r 4-feature-plots, cache=TRUE}
#Neurons:
list_plots <- lapply(c("Snap25", "Stmn2", "Rbfox3", "Gabrb2"),
                     function(x)plotTSNE(sce, colour_by = x ))

wrap_plots(list_plots) +  plot_annotation(title = "Neurons")
```

```{r violin-neuro, fig.height=4}
plotExpression(sce, features=c("Snap25", "Stmn2", "Rbfox3", "Gabrb2") ,
    x="cluster_k20", colour_by = "cluster_k20", ncol=1) +  scale_color_manual(values = unname(alphabet()))
```

```{r 4-feature-plots-default, cache=TRUE}
#Inhibitory neurons

list_plots <- lapply(c("Gad1", "Gad2", "Slc32a1", "Pvalb"),
                     function(x)plotTSNE(sce, colour_by = x ))
wrap_plots(list_plots) +  plot_annotation(title =  "Inhibitory Neurons")
```

```{r violin-neuroin, fig.height=4, cache=TRUE}
plotExpression(sce, features=c("Gad1", "Gad2", "Slc32a1", "Pvalb"),
    x="cluster_k20", colour_by = "cluster_k20", ncol=1) +  scale_color_manual(values = unname(alphabet()))
```

```{r one-line, fig.height=3, fig.width=9, cache=TRUE}
# Excitatory neurons

list_plots <- lapply(c("Satb2", "Slc12a6", "Slc17a7"),
                     function(x)plotTSNE(sce, colour_by = x )) 
wrap_plots(list_plots) +  plot_annotation(title =  "Exitatory Neurons")
```

```{r violin-neuroex, fig.height=3, cache=TRUE}
plotExpression(sce, features=c("Satb2", "Slc12a6", "Slc17a7"),
    x="cluster_k20", colour_by = "cluster_k20", ncol=1) +  scale_color_manual(values = unname(alphabet()))
```

```{r 4-plots, cache=TRUE}
# RBFOX3 (and other Granulate cell markers)
list_plots <- lapply(c("Cdh15", "Calb2", "Rbfox3", "Reln"),
function(x)plotTSNE(sce, colour_by = x )) 
wrap_plots(list_plots) +  plot_annotation(title =  "RBFO3+ Neurons")
```

```{r violin-neurorbfox, fig.height=4, cache=TRUE}
plotExpression(sce, features=c("Cdh15", "Calb2", "Rbfox3", "Reln"),
    x="cluster_k20", colour_by = "cluster_k20", ncol=1) +  scale_color_manual(values = unname(alphabet()))
```

</details>

<details>

<summary> Click to expand the Stromal marker plots </summary>

```{r small-feature-plots2, cache=TRUE}
#Stromal

list_plots <- lapply(c( "Lamb1" , 
                                     "Hspg2", 
                                     "Col4a1", 
                                     "Fn1", 
                                     "Lama2"),
            function(x) plotTSNE(sce, colour_by = x ))

wrap_plots(list_plots) +  plot_annotation(title =  "Stromal")
```

```{r violin-stromal, fig.height=5, cache=TRUE}
plotExpression(sce, features=c( "Lamb1" , 
                                     "Hspg2", 
                                     "Col4a1", 
                                     "Fn1", 
                                     "Lama2"),
    x="cluster_k20", colour_by = "cluster_k20", ncol=1) +  scale_color_manual(values = unname(alphabet()))
```

```{r 3lines-feature-plots, fig.height=8, cache=TRUE}
# Endothelial cells and pericytes

list_plots <- lapply(c( "Cldn5",   
                                     "Icam2",
                                     "Pdgfrb", 
                                     "Notch3", 
                                     "Vwf",
                                     "Flt1",
                                     "Mecom"),
                     function(x)plotTSNE(sce, colour_by = x ))
wrap_plots(list_plots) +  plot_annotation(title =  " Endothelial cells and pericytes")
```

```{r violin-endo, fig.height=7, cache=TRUE}
plotExpression(sce, features=c( "Cldn5",   
                                     "Icam2",
                                     "Pdgfrb", 
                                     "Notch3", 
                                     "Vwf",
                                     "Flt1",
                                     "Mecom"),
    x="cluster_k20", colour_by = "cluster_k20", ncol=1) +  scale_color_manual(values = unname(alphabet()))
```

</details>

<details>

<summary>Click to expand the Epithelial cells </summary>

```{r one-line-3-col, fig.height=3, fig.width=9, cache=TRUE}
list_plots <- lapply(c("Ttr", "Kcnj13", "Krt18"),
                     function(x)plotTSNE(sce, colour_by = x ))
wrap_plots(list_plots) +  plot_annotation(title =  "Choroid plexus epithelial cells")
```

```{r violin-epi, fig.height=3, cache=TRUE}
plotExpression(sce, features=c("Ttr", "Kcnj13", "Krt18"),
    x="cluster_k20", colour_by = "cluster_k20", ncol=1) +  scale_color_manual(values = unname(alphabet()))
```

</details>

<details>

<summary> Click to expand the Astrocytes marker plots </summary>

```{r 3lines-4-col-feature-plots, fig.height=7, fig.width=9, cache=TRUE}
# Astrocytes

list_plots <- lapply(c("Gja1",
                                    "Aqp4", 
                                    "Glul", 
                                    "Sox9", 
                                    "Ndrg2", 
                                    "Gfap", 
                                    "Aldh1a1", 
                                    "Aldh1l1", 
                                    "Vim", 
                                    "Apoe", 
                                    "Fgfr3"),
                     function(x)plotTSNE(sce, colour_by = x ))
wrap_plots(list_plots) +  plot_annotation(title = "Astrocyte")
```

```{r violin-astro, fig.height=11, cache=TRUE}
plotExpression(sce, features=c("Gja1",
                                    "Aqp4", 
                                    "Glul", 
                                    "Sox9", 
                                    "Ndrg2", 
                                    "Gfap", 
                                    "Aldh1a1", 
                                    "Aldh1l1", 
                                    "Vim", 
                                    "Apoe", 
                                    "Fgfr3"),
    x="cluster_k20", colour_by = "cluster_k20", ncol=1) +  scale_color_manual(values = unname(alphabet()))
```

```{r 3lines-feature-plots-2, cache=TRUE, fig.height=9, fig.width=9}
#Astrocyte markers as described in Zeisel et al. 2018 in the mouse for telencephalon and non-telencephalon astrocytes

list_plots <- lapply(c( "Agt", 
                                     "Mfge8",  
                                     "Slc6a11",
                                     "Slc6a9", 
                                     "Gdf10",
                                     "Islr",
                                     "Gfap",
                                     "Aqp4")  ,
                     function(x)plotTSNE(sce, colour_by = x ))
wrap_plots(list_plots) +  plot_annotation(title = "Astrocyte mouse telecephalon")
```

</details>

<details>

<summary> Click to expand the Immune cells plots </summary>

```{r 3lines-feature-plots3, cache=TRUE, fig.height=8}
#Microglia and macrophages

list_plots <- lapply(c( "Cd74", 
                                     "Spi1", 
                                     "Mrc1", 
                                     "Tmem119", 
                                     "Cx3cr1", 
                                     "Aif1",
                                     "P2ry12",
                                     "C1qc",
                                     "C1qa"),
            function(x)plotTSNE(sce, colour_by = x )) 
wrap_plots(list_plots) +  plot_annotation(title = "Microglia and macrophages")
```

```{r violin-micro, fig.height=9, cache=TRUE}
plotExpression(sce, features=c( "Cd74", 
                                     "Spi1", 
                                     "Mrc1", 
                                     "Tmem119", 
                                     "Cx3cr1", 
                                     "Aif1",
                                     "P2ry12",
                                     "C1qc",
                                     "C1qa"),
    x="cluster_k20", colour_by = "cluster_k20", ncol=1) +  scale_color_manual(values = unname(alphabet()))
```

```{r  3-in-one-line, fig.height=4, cache=TRUE}
# Border associated mcrophages
list_plots <- lapply(c( "Mrc1", "Ms4a7", "Apoe"),
            function(x)plotTSNE(sce, colour_by = x )) 
wrap_plots(list_plots) +  plot_annotation(title = "Border associated macrophages")
```

```{r violin-bam, fig.height=3, cache=TRUE}
plotExpression(sce, features=c( "Mrc1", "Ms4a7", "Apoe"),
    x="cluster_k20", colour_by = "cluster_k20", ncol=1) +  scale_color_manual(values = unname(alphabet()))
```

```{r  two-lines, fig.height=8, cache=TRUE}
# Immune cells 
list_plots <- lapply(c( 
  # Tcells
  "Cd3e", 
  # Bcells
  "Cd19", 
  # Natural killer
  "Klrb1c", "Cd209a", 
  # all immune cells (CD45)
  "Ptprc"),
            function(x)plotTSNE(sce, 
                                colour_by = x ,
                                point_alpha=0.3,
                                point_size = 0.5)) 
wrap_plots(list_plots) +  plot_annotation(title = "Immune cells")
```

```{r violin-immuno, fig.height=5, cache=TRUE}
plotExpression(sce, features=c( 
  # Tcells
  "Cd3e", 
  # Bcells
  "Cd19", 
  # Natural killer
  "Klrb1c", "Cd209a", 
  # all immune cells (CD45)
  "Ptprc"),
    x="cluster_k20", colour_by = "cluster_k20", ncol=1) +  scale_color_manual(values = unname(alphabet()))
```

```{r fig.height=4}
#monocybes/neutrophils
list_plots <- lapply(c( 
  # monocytes
  "S100a9",
  #neutrophils
  "Ly6g", "Camp"),
            function(x)plotTSNE(sce, 
                                colour_by = x ,
                                point_alpha=0.3,
                                point_size = 0.5)) 
wrap_plots(list_plots) +  plot_annotation(title = "Monocytes and Neutorphils cells")
```

```{r violin-mono, fig.height=3, cache=TRUE}
plotExpression(sce, features=c( 
   # monocytes
  "S100a9",
  #neutrophils
  "Ly6g", "Camp"),
    x="cluster_k20", colour_by = "cluster_k20", ncol=1) +  scale_color_manual(values = unname(alphabet()))
```

</details>

<details>

<summary> Click to expand the Oligodendroglia marker plots </summary>

```{r big-feature-plots4, cache=TRUE, fig.height=8}
#OPCs

list_plots <- lapply(c("Pdgfra", 
                                    "Cspg4", 
                                    "Gpr17", 
                                    "Ptprz1",
                                    "Olig1", 
                                    "Olig2", 
                                    "Pcdh15", 
                                    "Ptgds",
                                    "Bcan"),
            function(x)plotTSNE(sce, colour_by = x )) 
wrap_plots(list_plots) +  plot_annotation(title = "OPCS")
```

```{r violin-opc, fig.height=9, cache=TRUE}
plotExpression(sce, features=c("Pdgfra", 
                                    "Cspg4", 
                                    "Gpr17", 
                                    "Ptprz1",
                                    "Olig1", 
                                    "Olig2", 
                                    "Pcdh15", 
                                    "Ptgds",
                                    "Bcan"),
    x="cluster_k20", colour_by = "cluster_k20", ncol=1) +  scale_color_manual(values = unname(alphabet()))
```

```{r big-feature-plots5, cache=TRUE, fig.height=8}
#Oligodendrocytes

list_plots <- lapply(c("Plp1", 
                                    "Cnp", 
                                    "Mag", 
                                    "Mog", 
                                    "Mobp", 
                                    "Mbp", 
                                    "Sox10" ), 
            function(x)plotTSNE(sce, colour_by = x ))
wrap_plots(list_plots) +  plot_annotation(title =  "Oligodendrocytes")
```

```{r violin-oligo, fig.height=7, cache=TRUE}
plotExpression(sce, features=c("Plp1", 
                                    "Cnp", 
                                    "Mag", 
                                    "Mog", 
                                    "Mobp", 
                                    "Mbp", 
                                    "Sox10" ),
    x="cluster_k20", colour_by = "cluster_k20", ncol=1) +  scale_color_manual(values = unname(alphabet()))
```

</details>

<details>

<summary> Click to expand the ependymal marker plots </summary>

```{r ependymal, fig.height=8, fig.width=9, cache=TRUE}
list_plots <- lapply(c("Vit", "Sox9", "Dynlrb2", "Ccdc153", "Rsph1", "Tm4sf1", "Pcp4l1", "Pcp4", "Hspa2", "Cd24a", "Mt2", "Chchd10"), 
           function(x)plotTSNE(sce, colour_by = x )) 
wrap_plots(list_plots) +  plot_annotation(title =  "Ependymal")

```

</details>

## Top markers and differential expression

In order to help with the identification of we pull out the
most expressed genes for each cluster. ( ignoring the mitochondrial ribosomal genes)

```{r top_genes}
if(!file.exists(here("outs", age, "top_genes_k20_02.tsv"))){
# Create object where instead of cells in cols we have the clusters with counts aggregated
summed_k20 <- aggregateAcrossCells(sce, id=colData(sce)[,c("cluster_k20")])

# Delete ribosomal and mitochondrial genes
is_ribo <- grepl("^Rp[sl]", rownames(summed_k20))
is_mito <- grepl("^mt-", rownames(summed_k20))
keep <- !(is_ribo | is_mito)
 summed_k20 <- summed_k20[keep,]
 
# for each column (aggregated counts) of the object extract the top genes
top_genes_k20 <- apply(counts(summed_k20), 
      MARGIN = 2,
      function(x){
        # get the gene names from the top 50 expressed genes
        names(sort(x, decreasing = T)[1:100])
        }
      )
# save result
write.table(top_genes_k20, file = here("outs", age, "top_genes_k20_02.tsv"), sep = "\t", quote = FALSE, col.names = TRUE, row.names = FALSE)
}
```

We compute here a pairwise differential expression between all the clusters. The results are saved in a list of dataframes, "markers_k20_02/markers_k20_02.RDS"
one for each cluster. A .csv file (that can be opened with a spreadsheet program) for each cluster is available in "markers_k20_02".

The default philosophy of findMarkers() is to identify a combination of marker genes that - together - uniquely define one cluster against the rest. To this end, we collect the top DE genes from each pairwise comparison involving a particular cluster to assemble a set of candidate markers for that cluster. 
Of particular interest is the `Top` field. The set of genes with `Top` $≤X$ is the union of the top $X$ genes (ranked by p-value) from each pairwise comparison involving the cluster of interest. For example, the set of all genes with `Top` values of 1 contains the gene with the lowest p-value from each comparison. Similarly, the set of genes with `Top` values less than or equal to 10 contains the top 10 genes from each comparison.

```{r markers}
# compute markers
if(!file.exists(here("outs", age, "markers_k20_02", "markers_k20_02.RDS"))){
  dir.create(here("outs", age, "markers_k20_02"))
  markers <- findMarkers(sce, groups = sce$cluster_k20, direction="up", lfc=1)
  saveRDS(markers, here("outs", age,"markers_k20_02", "markers_k20_02.RDS"))
  # save the top 100 genes for each one of the df
  lapply(names(markers), function(x){
    top_markers <- head(markers[[x]], 100)
    write.csv(top_markers, here("outs", age, "markers_k20_02", paste0("top_markers_", x, "k20_02.csv")), quote = FALSE)
    }
  )
}
```

## Rename the Clusters with assigned Cell Types

With the help of the known markers plotted above, the DE between clusters and the top 
expressed genes the celltype identity of every cluster has benn identified. 

Renaming the clusters accordingly:

1 mNeurons

2 Endothelial

3 fEndothelia

4 Gran & Mono

5 OligoAstro

6 Microglia

7 Mural_cells

8 Lymphocytes

9 ChP_epithelial

10 OPCs

11 Astrocyte_1

12 Oligo_2

13 Astrocyte_2

14 Astrocyte_3

15 Oligo_1

16 DCs

17 Microglia

18 iNeuron & NRPs

19 BAMs

20 Ependymocytes

21 Microglia

22 OEG


```{r}
plotTSNE(sce, colour_by = "cluster_k20", text_by = "cluster_k20") + scale_color_manual(values = unname(alphabet()))
```

```{r result}
if (!file.exists(here("processed", age, "sce_anno_02.RDS"))) {
sce$celltype <- sce$cluster_k20 
# add the celltypes as described above in the correct order to replace the levels.
levels(sce$celltype) <- c("Neuron","Endothelial","Endothelial","Gran & Mono","OligoAstro","Microglia","Mural_cells","Lymphocytes","ChP_epithelial","OPCs","Astrocyte","Oligo","Astrocyte","Astrocyte","Oligo","DCs","Microglia","Neuron","BAMs","Ependymocytes","Microglia","OEG")
# and the clusters named to keep the resolution
sce$clusters_named <- sce$cluster_k20
levels(sce$clusters_named) <- c("mNeurons", "Endothelial", "fEndothelia", "Gran & Mono", "OligoAstro", "Microglia", "Mural_cells", "Lymphocytes", "ChP_epithelial", "OPCs", "Astrocyte_1", "Oligo_2", "Astrocyte_2", "Astrocyte_3", "Oligo_1", "DCs", "Microglia", "iNeurons & NRPs", "BAMs", "Ependymocytes", "Microglia", "OEG")

# relevel to get the appropriate order in the plots
sce$genotype <- relevel(factor(sce$genotype), "WT")
sce$clusters_named <- 
  factor(sce$clusters_named, 
         levels = c("Astrocyte_1", "Astrocyte_2", "Astrocyte_3", "OligoAstro", "Oligo_1", "Oligo_2", "OPCs", "mNeurons", "iNeurons & NRPs", "Lymphocytes", "Gran & Mono", "BAMs", "Microglia", "DCs",  "Endothelial", "fEndothelia", "Mural_cells", "ChP_epithelial", "Ependymocytes", "OEG"))

sce$celltype <- 
  factor(sce$celltype, 
         levels = c("Astrocyte", "OligoAstro", "Oligo", "OPCs", "Neuron", "Lymphocytes", "Gran & Mono", "BAMs", "Microglia", "DCs",  "Endothelial", "Mural_cells", "ChP_epithelial", "Ependymocytes", "OEG"))

 saveRDS(sce, here("processed", age, "sce_anno_02.RDS"))
}
plotTSNE(sce, colour_by = "clusters_named", text_by = "clusters_named", text_size = 3) + scale_color_manual(values = kelly_col)
```

## Marker plots

Exploring the shiny app and the DE between clusters specific markers for each one of the subtypes is identified.

```{r final-violin, fig.height=25, fig.width=9, cache=TRUE}
plotExpression(sce, features=c(# astrocytes
                                 "Aqp4",
                                 #astro1
                                 "Pantr1",
                                # astro 2
                                "Agt",
                                # astro 3
                                "Dclk1",
                                # oligos
                                "Mog",
                                #oligo1
                                "Ptgds",
                                #oligo2
                                "S100b",
                                #opcs
                                "Pdgfra",
                                #neurons
                                "Syt1",
                                #ineurons
                                "Sox11",
                                #lympho
                                "Ptprcap",
                                # granmono
                                "S100a9",
                                # Bams
                                "Mrc1",
                                # Microglia
                                "P2ry12",
                                # DCs
                                "Flt3",
                                # endothelial
                                "Pecam1",
                                # fendothelial
                                "Plvap",
                                # mural cells
                                "Vtn",
                                #chp
                                "Ttr",
                                #ependymocytes
                                "Ccdc153",
                                #oeg
                                "Npy"
                                
                                ),
    x="clusters_named", colour_by = "clusters_named", ncol=1, scales = "free_y") +  
  scale_color_manual(values = kelly_col) + theme(axis.text.x = element_text(angle = 45, hjust=1, vjust = 1)) 

```

## Session Info

<details>

<summary> Click to expand </summary>

```{r session-info}
sessionInfo()
```

</details>
