---
title: "Annotation"
author: "NadineBestard"
date: "30/03/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE, warning=FALSE}
library(here) #reproducible paths
library(scater) #feature plots
library(patchwork) # agregate plots
library(scran) # for findmarkers
```

```{r age}
age <- "old"
```

```{r import}
if (!file.exists(here("processed", age, "sce_anno_01.RDS"))) {
  sce <- readRDS(here("processed", age, "sce_clusters_01.RDS"))
}else{
  sce <- readRDS(here("processed", age, "sce_anno_01.RDS"))
}
```

## Annotation

We make a first rough annotation using known markers for different celltypes. 

<details>
  <summary>Click to expand the Neurons marker plots</summary>
```{r 4-feature-plots, cache=TRUE}
#Neurons:
list_plots <- lapply(c("Snap25", "Stmn2", "Rbfox3", "Gabrb2"),
                     function(x)plotTSNE(sce, colour_by = x ))

wrap_plots(list_plots) +  plot_annotation(title = "Neurons")

#Inhibitory neurons

list_plots <- lapply(c("Gad1", "Gad2", "Slc32a1", "Pvalb"),
                     function(x)plotTSNE(sce, colour_by = x ))
wrap_plots(list_plots) +  plot_annotation(title =  "Inhibitory Neurons")
```


```{r one-line, fig.height=4, cache=TRUE}
# Excitatory neurons

list_plots <- lapply(c("Satb2", "Slc12a6", "Slc17a7"),
                     function(x)plotTSNE(sce, colour_by = x )) 
wrap_plots(list_plots) +  plot_annotation(title =  "Exitatory Neurons")
```


```{r 4-plots, cache=TRUE}
# RBFOX3 (and other Granulate cell markers)
list_plots <- lapply(c("Cdh15", "Calb2", "Rbfox3", "Reln"),
function(x)plotTSNE(sce, colour_by = x )) 
wrap_plots(list_plots) +  plot_annotation(title =  "RBFO3+ Neurons")
```

</details>
<details>
  <summary>Click to expand the Stromal  marker plots</summary>
```{r small-feature-plots2, cache=TRUE}
#Stromal

list_plots <- lapply(c( "Lamb1" , 
                                     "Hspg2", 
                                     "Col4a1", 
                                     "Fn1", 
                                     "Lama2"),
            function(x) plotTSNE(sce, colour_by = x ))

wrap_plots(list_plots) +  plot_annotation(title =  "Stromal")
```




```{r 3lines-feature-plots, cache=TRUE, fig.height=8}
# Endothelial cells and pericytes

list_plots <- lapply(c( "Cldn5",   
                                     "Icam2",
                                     "Pdgfrb", 
                                     "Notch3", 
                                     "Vwf",
                                     "Flt1",
                                     "Mecom"),
                     function(x)plotTSNE(sce, colour_by = x ))
wrap_plots(list_plots) +  plot_annotation(title =  " Endothelial cells and pericytes")
```
</details>

<details>
  <summary>Click to expand the Epithelial cells</summary>
```{r one-line-3-col, fig.height=4, cache=TRUE}
list_plots <- lapply(c("Ttr", "Kcnj13", "Krt18"),
                     function(x)plotTSNE(sce, colour_by = x ))
wrap_plots(list_plots) +  plot_annotation(title =  "Choroid plexus epithelial cells")
```

</details>


<details>
  <summary>Click to expand the Astrocytes marker plots</summary>
```{r 3lines-4-col-feature-plots, fig.height=8, fig.width=9, cache=TRUE}
# Astrocytes

list_plots <- lapply(c("Gja1",
                                    "Aqp4", 
                                    "Glul", 
                                    "Sox9", 
                                    "Ndrg2", 
                                    "Gfap", 
                                    "Aldh1a1", 
                                    "Aldh1l1", 
                                    "Vim", 
                                    "Apoe", 
                                    "Fgfr3"),
                     function(x)plotTSNE(sce, colour_by = x ))
wrap_plots(list_plots) +  plot_annotation(title = "Astrocyte")
```


```{r 3lines-feature-plots-2, cache=TRUE, fig.height=8}
#Astrocyte markers as described in Zeisel et al. 2018 in the mouse for telencephalon and non-telencephalon astrocytes

list_plots <- lapply(c( "Agt", 
                                     "Mfge8",  
                                     "Slc6a11",
                                     "Slc6a9", 
                                     "Gdf10",
                                     "Islr",
                                     "Gfap",
                                     "Aqp4")  ,
                     function(x)plotTSNE(sce, colour_by = x ))
wrap_plots(list_plots) +  plot_annotation(title = "Astrocyte mouse telecephalon")
```

</details>

<details>
  <summary>Click to expand the Immune cells plots</summary>

```{r 3lines-feature-plots3, cache=TRUE, fig.height=8}
#Microglia and macrophages

list_plots <- lapply(c( "Cd74", 
                                     "Spi1", 
                                     "Mrc1", 
                                     "Tmem119", 
                                     "Cx3cr1", 
                                     "Aif1",
                                     "P2ry12",
                                     "C1qc",
                                     "C1qa"),
            function(x)plotTSNE(sce, colour_by = x )) 
wrap_plots(list_plots) +  plot_annotation(title = "Microglia and macrophages")
```

```{r  3-in-one-line, fig.height=4, cache=TRUE}
# Border associated mcrophages
list_plots <- lapply(c( "Mrc1", "Ms4a7", "Apoe"),
            function(x)plotTSNE(sce, colour_by = x )) 
wrap_plots(list_plots) +  plot_annotation(title = "Border associated macrophages")
```

```{r  two-lines, fig.height=8, cache=TRUE}
# Immune cells 
list_plots <- lapply(c( 
  # Tcells
  "Cd3e", 
  # Bcells
  "Cd19", 
  # Natural killer
  "Klrb1c", "Cd209a", 
  # all immune cells (CD45)
  "Ptprc"),
            function(x)plotTSNE(sce, 
                                colour_by = x ,
                                point_alpha=0.3,
                                point_size = 0.5)) 
wrap_plots(list_plots) +  plot_annotation(title = "Immune cells")
```

```{r fig.height=4}
#monocybes/neutrophils
list_plots <- lapply(c( 
  # monocytes
  "S100a9",
  #neutrophils
  "Ly6g", "Camp"),
            function(x)plotTSNE(sce, 
                                colour_by = x ,
                                point_alpha=0.3,
                                point_size = 0.5)) 
wrap_plots(list_plots) +  plot_annotation(title = "Monocytes and Neutorphils cells")
```

</details>
<details>
  <summary>Click to expand the Oligodendroglia marker plots</summary>
```{r big-feature-plots4, cache=TRUE, fig.height=8}
#OPCs

list_plots <- lapply(c("Pdgfra", 
                                    "Cspg4", 
                                    "Gpr17", 
                                    "Ptprz1",
                                    "Olig1", 
                                    "Olig2", 
                                    "Pcdh15", 
                                    "Ptgds",
                                    "Bcan"),
            function(x)plotTSNE(sce, colour_by = x )) 
wrap_plots(list_plots) +  plot_annotation(title = "OPCS")
            
#Oligodendrocytes

list_plots <- lapply(c("Plp1", 
                                    "Cnp", 
                                    "Mag", 
                                    "Mog", 
                                    "Mobp", 
                                    "Mbp", 
                                    "Sox10" ), 
            function(x)plotTSNE(sce, colour_by = x ))
wrap_plots(list_plots) +  plot_annotation(title =  "Oligodendrocytes")
```

</details>
<details>
  <summary>Click to expand the ependymal marker plots</summary>
```{r ependymal, fig.height=8, fig.width=9, cache=TRUE}
list_plots <- lapply(c("Vit", "Sox9", "Dynlrb2", "Ccdc153", "Rsph1", "Tm4sf1", "Pcp4l1", "Pcp4", "Hspa2", "Cd24a", "Mt2", "Chchd10"), 
           function(x)plotTSNE(sce, colour_by = x )) 
wrap_plots(list_plots) +  plot_annotation(title =  "Ependymal")

```
  
</details>



## Extra. identify cluster 10

In order to help with the identification of cluster 10 we pull out the most expressed genes, 
and the DE with the other clusters.

```{r}
# subset only the cluster of interest
sce_10 <- sce[ ,sce$cluster_k60 == 10]
# calculate the average expression for each gene and extract the top 50
avg_10 <- rowSums(counts(sce_10))/ncol(counts(sce_10))
top_genes_10 <- sort(avg_10, decreasing = TRUE)[1:50]
# save result
write.table(top_genes_10, file = here("outs", age, "top_genes_10k60_01.tsv"), sep = "\t", quote = FALSE, col.names = FALSE, row.names = TRUE)
```
The top expressed genes are mostly ribosomal genes, there are often very highly expressed genes, and not very informative.

We perform a DE between cluster 10 and the other clusters to pull out more interesting genes from this cluster, this also yields ribosomal genes and genes that do not seem specific to any particular celltype. We will have a closer look in the cluster QC.
```{r}
# compute markers
if(!file.exists(here("outs", age, "markers_k60_01.RDS"))){
  markers <- findMarkers(sce, groups = sce$cluster_k60, direction="up", lfc=1)
  saveRDS(markers, here("outs", age, "markers_k60_01.RDS"))
}else{
  markers <- readRDS(here("outs", age, "markers_k60_01.RDS"))
}
# select cluster 10
markers_10 <- markers[["10"]]
top_markers <- head(markers_10, 50)
write.csv(top_markers, here("outs", age, "top_markers_10k60_01.csv"), quote = FALSE)
```


## Rename the Clusters with assigned Cell Types

For this first annotation we will rename the levels from the k=60 clustering as 
follows:
1 	Microglia-Macrophages
2	Endothelial-Pericytes
3	Astrocytes
4	OPC
5	Immune
6	Astrocytes
7	Monocytes
8	Astrocytes
9 	Oligo
10	NA
11	Chp-epithelial
12	Oligo
13	NA
14	Oligo
15	Neuron
16 	Microglia-Macrophages-BAM
17	Oligo?
18	Oligo
19	OPC
20	Endothelial-Pericytes
21 	Microglia-Macrophages
22	Endothelial-Pericytes
23	Astrocytes
24 	Microglia-Macrophages

```{r}
plotTSNE(sce, colour_by = "cluster_k60", text_by = "cluster_k60") + scale_color_hue()
```

```{r result}
if (!file.exists(here("processed", age, "sce_anno_01.RDS"))) {
sce$celltype <- sce$cluster_k60 
# add the celltypes as described above in the correct order to replace the levels.
levels(sce$celltype)<-c("Microglia-Macrophages", "Endothelial-Pericytes", "Astrocytes", "OPC", "Immune", "Astrocytes", "Monocytes", "Astrocytes", "Oligo", "NA", "ChP-epithelial", "Oligo", "NA", "Oligo", "Neuron", "Microglia-Macrophages-BAM", "Oligo?", "Oligo", "OPC", "Endothelial-Pericytes", "Microglia-Macrophages", "Endothelial-Pericytes", "Astrocytes", "Microglia-Macrophages")
 saveRDS(sce, here("processed", age, "sce_anno_01.RDS"))
}
plotTSNE(sce, colour_by = "celltype", text_by = "cluster_k60") + scale_color_hue()

```

