---
title: "Annotation cluster k20"
author: "NadineBestard"
date: "30/03/2021"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE, warning=FALSE}
library(here) #reproducible paths
library(scater) #feature plots
library(patchwork) # agregate plots
library(scran) # for findmarkers
```

```{r age}
age <- "old"
```

```{r import}
if (!file.exists(here("processed", age, "sce_anno_02.RDS"))) {
  sce <- readRDS(here("processed", age, "sce_clusters_02.RDS"))
}else{
  sce <- readRDS(here("processed", age, "sce_anno_02.RDS"))
}
```

## Annotation

We make a first rough annotation using known markers for different
celltypes.


```{r}
plotTSNE(sce, colour_by = "cluster_k20", text_by = "cluster_k20") + scale_color_hue()
```

<details>

<summary> Click to expand the Neurons marker plots </summary>

```{r 4-feature-plots, cache=TRUE}
#Neurons:
list_plots <- lapply(c("Snap25", "Stmn2", "Rbfox3", "Gabrb2"),
                     function(x)plotTSNE(sce, colour_by = x ))

wrap_plots(list_plots) +  plot_annotation(title = "Neurons")
```

```{r violin-neuro, fig.height=4}
plotExpression(sce, features=c("Snap25", "Stmn2", "Rbfox3", "Gabrb2") ,
    x="cluster_k20", colour_by = "cluster_k20", ncol=1) + scale_color_hue()
```

```{r 4-feature-plots-default, cache=TRUE}
#Inhibitory neurons

list_plots <- lapply(c("Gad1", "Gad2", "Slc32a1", "Pvalb"),
                     function(x)plotTSNE(sce, colour_by = x ))
wrap_plots(list_plots) +  plot_annotation(title =  "Inhibitory Neurons")
```

```{r violin-neuroin, fig.height=4, cache=TRUE}
plotExpression(sce, features=c("Gad1", "Gad2", "Slc32a1", "Pvalb"),
    x="cluster_k20", colour_by = "cluster_k20", ncol=1) + scale_color_hue()
```

```{r one-line, fig.height=3, fig.width=9, cache=TRUE}
# Excitatory neurons

list_plots <- lapply(c("Satb2", "Slc12a6", "Slc17a7"),
                     function(x)plotTSNE(sce, colour_by = x )) 
wrap_plots(list_plots) +  plot_annotation(title =  "Exitatory Neurons")
```

```{r violin-neuroex, fig.height=3, cache=TRUE}
plotExpression(sce, features=c("Satb2", "Slc12a6", "Slc17a7"),
    x="cluster_k20", colour_by = "cluster_k20", ncol=1) + scale_color_hue()
```

```{r 4-plots, cache=TRUE}
# RBFOX3 (and other Granulate cell markers)
list_plots <- lapply(c("Cdh15", "Calb2", "Rbfox3", "Reln"),
function(x)plotTSNE(sce, colour_by = x )) 
wrap_plots(list_plots) +  plot_annotation(title =  "RBFO3+ Neurons")
```

```{r violin-neurorbfox, fig.height=4, cache=TRUE}
plotExpression(sce, features=c("Cdh15", "Calb2", "Rbfox3", "Reln"),
    x="cluster_k20", colour_by = "cluster_k20", ncol=1) + scale_color_hue()
```

</details>

<details>

<summary> Click to expand the Stromal marker plots </summary>

```{r small-feature-plots2, cache=TRUE}
#Stromal

list_plots <- lapply(c( "Lamb1" , 
                                     "Hspg2", 
                                     "Col4a1", 
                                     "Fn1", 
                                     "Lama2"),
            function(x) plotTSNE(sce, colour_by = x ))

wrap_plots(list_plots) +  plot_annotation(title =  "Stromal")
```

```{r violin-stromal, fig.height=5, cache=TRUE}
plotExpression(sce, features=c( "Lamb1" , 
                                     "Hspg2", 
                                     "Col4a1", 
                                     "Fn1", 
                                     "Lama2"),
    x="cluster_k20", colour_by = "cluster_k20", ncol=1) + scale_color_hue()
```

```{r 3lines-feature-plots, fig.height=8, cache=TRUE}
# Endothelial cells and pericytes

list_plots <- lapply(c( "Cldn5",   
                                     "Icam2",
                                     "Pdgfrb", 
                                     "Notch3", 
                                     "Vwf",
                                     "Flt1",
                                     "Mecom"),
                     function(x)plotTSNE(sce, colour_by = x ))
wrap_plots(list_plots) +  plot_annotation(title =  " Endothelial cells and pericytes")
```

```{r violin-endo, fig.height=7, cache=TRUE}
plotExpression(sce, features=c( "Cldn5",   
                                     "Icam2",
                                     "Pdgfrb", 
                                     "Notch3", 
                                     "Vwf",
                                     "Flt1",
                                     "Mecom"),
    x="cluster_k20", colour_by = "cluster_k20", ncol=1) + scale_color_hue()
```

</details>

<details>

<summary>Click to expand the Epithelial cells </summary>

```{r one-line-3-col, fig.height=3, fig.width=9, cache=TRUE}
list_plots <- lapply(c("Ttr", "Kcnj13", "Krt18"),
                     function(x)plotTSNE(sce, colour_by = x ))
wrap_plots(list_plots) +  plot_annotation(title =  "Choroid plexus epithelial cells")
```

```{r violin-epi, fig.height=3, cache=TRUE}
plotExpression(sce, features=c("Ttr", "Kcnj13", "Krt18"),
    x="cluster_k20", colour_by = "cluster_k20", ncol=1) + scale_color_hue()
```

</details>

<details>

<summary>

Click to expand the Astrocytes marker plots

</summary>

```{r 3lines-4-col-feature-plots, fig.height=7, fig.width=9, cache=TRUE}
# Astrocytes

list_plots <- lapply(c("Gja1",
                                    "Aqp4", 
                                    "Glul", 
                                    "Sox9", 
                                    "Ndrg2", 
                                    "Gfap", 
                                    "Aldh1a1", 
                                    "Aldh1l1", 
                                    "Vim", 
                                    "Apoe", 
                                    "Fgfr3"),
                     function(x)plotTSNE(sce, colour_by = x ))
wrap_plots(list_plots) +  plot_annotation(title = "Astrocyte")
```

```{r violin-astro, fig.height=11, cache=TRUE}
plotExpression(sce, features=c("Gja1",
                                    "Aqp4", 
                                    "Glul", 
                                    "Sox9", 
                                    "Ndrg2", 
                                    "Gfap", 
                                    "Aldh1a1", 
                                    "Aldh1l1", 
                                    "Vim", 
                                    "Apoe", 
                                    "Fgfr3"),
    x="cluster_k20", colour_by = "cluster_k20", ncol=1) + scale_color_hue()
```

```{r 3lines-feature-plots-2, cache=TRUE, fig.height=9, fig.width=9}
#Astrocyte markers as described in Zeisel et al. 2018 in the mouse for telencephalon and non-telencephalon astrocytes

list_plots <- lapply(c( "Agt", 
                                     "Mfge8",  
                                     "Slc6a11",
                                     "Slc6a9", 
                                     "Gdf10",
                                     "Islr",
                                     "Gfap",
                                     "Aqp4")  ,
                     function(x)plotTSNE(sce, colour_by = x ))
wrap_plots(list_plots) +  plot_annotation(title = "Astrocyte mouse telecephalon")
```

</details>

<details>

<summary> Click to expand the Immune cells plots </summary>

```{r 3lines-feature-plots3, cache=TRUE, fig.height=8}
#Microglia and macrophages

list_plots <- lapply(c( "Cd74", 
                                     "Spi1", 
                                     "Mrc1", 
                                     "Tmem119", 
                                     "Cx3cr1", 
                                     "Aif1",
                                     "P2ry12",
                                     "C1qc",
                                     "C1qa"),
            function(x)plotTSNE(sce, colour_by = x )) 
wrap_plots(list_plots) +  plot_annotation(title = "Microglia and macrophages")
```

```{r violin-micro, fig.height=9, cache=TRUE}
plotExpression(sce, features=c( "Cd74", 
                                     "Spi1", 
                                     "Mrc1", 
                                     "Tmem119", 
                                     "Cx3cr1", 
                                     "Aif1",
                                     "P2ry12",
                                     "C1qc",
                                     "C1qa"),
    x="cluster_k20", colour_by = "cluster_k20", ncol=1) + scale_color_hue()
```

```{r  3-in-one-line, fig.height=4, cache=TRUE}
# Border associated mcrophages
list_plots <- lapply(c( "Mrc1", "Ms4a7", "Apoe"),
            function(x)plotTSNE(sce, colour_by = x )) 
wrap_plots(list_plots) +  plot_annotation(title = "Border associated macrophages")
```

```{r violin-bam, fig.height=3, cache=TRUE}
plotExpression(sce, features=c( "Mrc1", "Ms4a7", "Apoe"),
    x="cluster_k20", colour_by = "cluster_k20", ncol=1) + scale_color_hue()
```

```{r  two-lines, fig.height=8, cache=TRUE}
# Immune cells 
list_plots <- lapply(c( 
  # Tcells
  "Cd3e", 
  # Bcells
  "Cd19", 
  # Natural killer
  "Klrb1c", "Cd209a", 
  # all immune cells (CD45)
  "Ptprc"),
            function(x)plotTSNE(sce, 
                                colour_by = x ,
                                point_alpha=0.3,
                                point_size = 0.5)) 
wrap_plots(list_plots) +  plot_annotation(title = "Immune cells")
```

```{r violin-immuno, fig.height=5, cache=TRUE}
plotExpression(sce, features=c( 
  # Tcells
  "Cd3e", 
  # Bcells
  "Cd19", 
  # Natural killer
  "Klrb1c", "Cd209a", 
  # all immune cells (CD45)
  "Ptprc"),
    x="cluster_k20", colour_by = "cluster_k20", ncol=1) + scale_color_hue()
```

```{r fig.height=4}
#monocybes/neutrophils
list_plots <- lapply(c( 
  # monocytes
  "S100a9",
  #neutrophils
  "Ly6g", "Camp"),
            function(x)plotTSNE(sce, 
                                colour_by = x ,
                                point_alpha=0.3,
                                point_size = 0.5)) 
wrap_plots(list_plots) +  plot_annotation(title = "Monocytes and Neutorphils cells")
```

```{r violin-mono, fig.height=3, cache=TRUE}
plotExpression(sce, features=c( 
   # monocytes
  "S100a9",
  #neutrophils
  "Ly6g", "Camp"),
    x="cluster_k20", colour_by = "cluster_k20", ncol=1) + scale_color_hue()
```

</details>

<details>

<summary> Click to expand the Oligodendroglia marker plots </summary>

```{r big-feature-plots4, cache=TRUE, fig.height=8}
#OPCs

list_plots <- lapply(c("Pdgfra", 
                                    "Cspg4", 
                                    "Gpr17", 
                                    "Ptprz1",
                                    "Olig1", 
                                    "Olig2", 
                                    "Pcdh15", 
                                    "Ptgds",
                                    "Bcan"),
            function(x)plotTSNE(sce, colour_by = x )) 
wrap_plots(list_plots) +  plot_annotation(title = "OPCS")
```

```{r violin-opc, fig.height=9, cache=TRUE}
plotExpression(sce, features=c("Pdgfra", 
                                    "Cspg4", 
                                    "Gpr17", 
                                    "Ptprz1",
                                    "Olig1", 
                                    "Olig2", 
                                    "Pcdh15", 
                                    "Ptgds",
                                    "Bcan"),
    x="cluster_k20", colour_by = "cluster_k20", ncol=1) + scale_color_hue()
```

```{r big-feature-plots5, cache=TRUE, fig.height=8}
#Oligodendrocytes

list_plots <- lapply(c("Plp1", 
                                    "Cnp", 
                                    "Mag", 
                                    "Mog", 
                                    "Mobp", 
                                    "Mbp", 
                                    "Sox10" ), 
            function(x)plotTSNE(sce, colour_by = x ))
wrap_plots(list_plots) +  plot_annotation(title =  "Oligodendrocytes")
```

```{r violin-oligo, fig.height=7, cache=TRUE}
plotExpression(sce, features=c("Plp1", 
                                    "Cnp", 
                                    "Mag", 
                                    "Mog", 
                                    "Mobp", 
                                    "Mbp", 
                                    "Sox10" ),
    x="cluster_k20", colour_by = "cluster_k20", ncol=1) + scale_color_hue()
```

</details>

<details>

<summary> Click to expand the ependymal marker plots </summary>

```{r ependymal, fig.height=8, fig.width=9, cache=TRUE}
list_plots <- lapply(c("Vit", "Sox9", "Dynlrb2", "Ccdc153", "Rsph1", "Tm4sf1", "Pcp4l1", "Pcp4", "Hspa2", "Cd24a", "Mt2", "Chchd10"), 
           function(x)plotTSNE(sce, colour_by = x )) 
wrap_plots(list_plots) +  plot_annotation(title =  "Ependymal")

```

</details>

## Top markers and differential expression

In order to help with the identification of we pull out the
most expressed genes for each cluster. ( ignoring the mithocondrial ribosomal, apoe, malat1)

```{r top_genes}
if(!file.exists(here("outs", age, "top_genes_k20_02.tsv"))){
# Create object where instead of cells in cols we have the clusters with counts aggregated
summed_k20 <- aggregateAcrossCells(sce, id=colData(sce)[,c("cluster_k20")])

# Delete very expressed/not informative genes such as ribosomal and mitocondrial genes
is_ribo <- grepl("^Rp[sl]", rownames(summed_k20))
is_mito <- grepl("^mt-", rownames(summed_k20))
is_abundant <- rownames(summed_k20) %in% c("Malat1", "Apoe")
keep <- !(is_ribo | is_mito|is_abundant)
 summed_k20 <- summed_k20[keep,]
 
# for each column (aggregated counts) of the object extract the top genes
top_genes_k20 <- apply(counts(summed_k20), 
      MARGIN = 2,
      function(x){
        # get the gene names from the top 50 expressed genes
        names(sort(x, decreasing = T)[1:100])
        }
      )
# save result
write.table(top_genes_k20, file = here("outs", age, "top_genes_k20_02.tsv"), sep = "\t", quote = FALSE, col.names = TRUE, row.names = FALSE)
}
```

We compute here a pairwise differential expression between all the clusters. The results are saved in a list of dataframes, "markers_k20_02/markers_k20_02.RDS"
one for each cluster. A .csv file (that can be opened with a spreadsheet program) for each cluster is available in "markers_k20_02".

The default philosophy of findMarkers() is to identify a combination of marker genes that - together - uniquely define one cluster against the rest. To this end, we collect the top DE genes from each pairwise comparison involving a particular cluster to assemble a set of candidate markers for that cluster. 
Of particular interest is the `Top` field. The set of genes with `Top` $≤X$ is the union of the top $X$ genes (ranked by p-value) from each pairwise comparison involving the cluster of interest. For example, the set of all genes with `Top` values of 1 contains the gene with the lowest p-value from each comparison. Similarly, the set of genes with `Top` values less than or equal to 10 contains the top 10 genes from each comparison.

```{r markers}
# compute markers
if(!file.exists(here("outs", age, "markers_k20_02", "markers_k20_02.RDS"))){
  dir.create(here("outs", age, "markers_k20_02"))
  markers <- findMarkers(sce, groups = sce$cluster_k20, direction="up", lfc=1)
  saveRDS(markers, here("outs", age,"markers_k20_02", "markers_k20_02.RDS"))
  # save the top 100 genes for each one of the df
  lapply(names(markers), function(x){
    top_markers <- head(markers[[x]], 100)
    write.csv(top_markers, here("outs", age, "markers_k20_02", paste0("top_markers_", x, "k20_02.csv")), quote = FALSE)
    }
  )
}
```

## Rename the Clusters with assigned Cell Types

For this first annotation we will rename the levels from the k=20
clustering as follows:

1 Neuron, This splits at k10

2 Endothelial-Pericytes

3 Endothelial-Pericytes, with some RELN+, should I go higher resolution?... 

4 Immune-Monocytes

5 Astrocyte-Oligo

6 Immune-Microglia-Macrophages. They could be distinguished at higher resolution

7 Endothelial-Pericytes

8 Immune

9 ChP-Epithelial

10 OPC

11 Astrocyte

12 Oligo

13 Astrocyte

14 Astrocyte

15 Oligo

16 Immune

17 Immune

18 Neuron

19 Immune-BAM

20 Pvalb-Astrocyte

21 Immune

22 Cldn5-Astrocyte-Plp1


```{r}
plotTSNE(sce, colour_by = "cluster_k20", text_by = "cluster_k20") + scale_color_hue()
```

```{r result}
if (!file.exists(here("processed", age, "sce_anno_02.RDS"))) {
sce$celltype <- sce$cluster_k20 
# add the celltypes as described above in the correct order to replace the levels.
levels(sce$celltype) <- c("Neuron","Endothelial-Pericytes","Endothelial-Pericytes","Immune-Monocytes","Astrocyte-Oligo","Immune-Microglia-Macrophages","Endothelial-Pericytes","Immune","ChP-Epithelial","OPC","Astrocyte","Oligo","Astrocyte","Astrocyte","Oligo","Immune","Immune","Neuron","Immune-BAM","Astrocyte-Pvalb","Immune","Astrocyte-Plp1-Cldn5")
# and the clusters named to keep the resolution
sce$cluster_k20_named <- sce$cluster_k20 
levels(sce$cluster_k20_named) <- c("Neuron_1","Endothelial-Pericytes_1","Endothelial-Pericytes_2","Immune-Monocytes","Astrocyte-Oligo","Immune-Microglia-Macrophages","Endothelial-Pericytes_3","Immune_1","ChP-Epithelial","OPC","Astrocyte_1","Oligo_2","Astrocyte_2","Astrocyte_3","Oligo_1","Immune_2","Immune_3","Neuron_2","Immune-BAM","Astrocyte-Pvalb","Immune_4","Astrocyte-Plp1-Cldn5")
 saveRDS(sce, here("processed", age, "sce_anno_02.RDS"))
}
plotTSNE(sce, colour_by = "celltype", text_by = "cluster_k20_named", text_size = 3) + scale_color_hue()
```


```{r eval=FALSE, include=FALSE}
plotReducedDimRepel(sce, dimred = "TSNE", colour_by = "celltype", text_by = "cluster_k20_named", text_size = 3) + scale_color_hue()
```
## Session Info

<details>

<summary> Click to expand </summary>

```{r session-info}
sessionInfo()
```

</details>
