---
title: "Cell Cycle"
author: "Nadine Bestard"
date: "04/08/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### set-up
```{r}
library(scRNAseq) # reference dataset
#library(scran) # for scDE
library(scater) # to manipulate sce
library(here) # reproducible paths
library(org.Mm.eg.db) # annotation cycling genes
library(SingleR) # to mapp cellcycle
library(DescTools) # save multidemensional arrays
```
```{r age, eval=T}
age <- "old"
```


```{r load}
sce <- readRDS(here("processed", age, "sce_anno_02.RDS"))
```
## Visualise cyclins

The cyclins control progression through the cell cycle and have well-characterized patterns of expression across cell cycle phases.
```{r fig.height=8}
ccn_genes <- rownames(sce)[grep("^Ccn[abde][0-9]$", rownames(sce))]

plotHeatmap(sce,
            columns =sce$genotype == "WT",
            order_columns_by=c("genotype","clusters_named"),
            colour_columns_by = c("genotype","clusters_named"),
            cluster_rows=FALSE, 
            cluster_cols = FALSE,
            features=ccn_genes) + facet_wrap(~ clusters_named)
```


## Use reference profile to perform cell cycle assignment

The key assumption is that the cell cycle effect is orthogonal to other aspects of biological heterogeneity like cell type. This justifies the use of a reference involving cell types that are different from the cells in our dataset, provided that the cell cycle transcriptional program is conserved across datasets (Bertoli, Skotheim, and Bruin 2013; Conboy et al. 2007).
Non-orthogonality can introduce biases where, one cell type is consistently misclassified as being in a particular phase because it happens to be more similar to that phase’s profile in the reference. 
A healthy dose of skepticism is required when interpreting these assignments. Our hope is that any systematic assignment error is consistent across clusters and conditions such that they cancel out in comparisons of phase frequencies.

```{r get-ref-profile}
# import and lognormalise
sce_ref <- BuettnerESCData()
sce_ref <- logNormCounts(sce_ref)
# filter for cycling genes
cycle_genes <- select(org.Mm.eg.db,
                           keytype = "GOALL",
                           keys = "GO:0007049",
                           columns = "ENSEMBL"
                           )[, "ENSEMBL"]
```

```{r singleR}
# from sce create matrix with only the logcounts and back with Ensembl notation
sce_mat <- logcounts(sce)
rownames(sce_mat) <- rowData(sce)$ID

# run assignments
cycle_cells <- SingleR(sce_mat, ref = sce_ref, labels = sce_ref$phase, de.method = "wilcox", restrict = cycle_genes)
```

```{r freqs}
# Extract frequencies and save to file.
# create a table with the frecuencies genotype-cellcycle for each one of the clusters
freq <- table(cycle=cycle_cells$pruned.labels,
              genotype=sce$genotype,
              cluster_named=sce$clusters_named)

# save to file
freq_flat <- stats:::format.ftable(ftable(freq, col.vars = c("genotype"), row.vars=c("cluster_named", "cycle")), quote=FALSE)
write.table(freq_flat, sep=",", file = here("outs", age, "cellcycle.csv"),row.names = FALSE, col.names = FALSE)
```

outs/`r age`/cellcycle.csv contains a table with the number of cells found in each phase, for each one of the clusters, divided by WT and KO.


## Run statistical test

In each one of our clusters we want to compare if the distribution of cells across the G1, G2M and S phases differs between the WT and the KO. Due to the presence of small counts (expected frequencies <5) we use the fisher's exact test, as advised in [(Hae-Young Kim et. al.)](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5426219/)


```{r test}
# Margin is 3 as I want to apply the test to each one of the clusters (3rd dimension)
apply(freq, 3, function(freq_WT.KO_G1.G2M.S){fisher.test(freq_WT.KO_G1.G2M.S)})

```


