---
title: "Cluster quality control"
author: "Nadine Bestard"
date: "28/05/2021"
output:
  html_document:
    df_print: paged
editor_options: 
  markdown: 
    wrap: 80
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

#### set-up

```{r set-up, message=FALSE, warning=FALSE}
library(SingleCellExperiment)
library(here) # reproducible paths
library(scater) # plot reduced dims
library(dplyr) # df filtering

```

```{r age}
age <- "old"
```

```{r load}
sce <- readRDS(here("processed", age, "sce_clusters.RDS"))
```

### Introduction

As in [publication] we will perform a cluster QC to remove clusters of poorer
quality. This will be assessed by the number of UMI counts, the mitochondrial
percentage and the number of mice that contribute to each cluster. To do so we
use a small cluster resolution, 5

```{r}
plotTSNE(sce, colour_by = "cluster_k5", point_size=0.5,
         point_alpha = 0.3, text_by = "cluster_k5", text_size = 3) +
  scale_color_hue() 
```

### Number of molecules per cluster

```{r}
plotTSNE(sce, colour_by = "total") + 
  ggtitle("Total number of umi counts per cell")
plotTSNE(sce, colour_by = "detected") +
  ggtitle("Detected number of genes per cell")
```

Lower values of umi counts and detected genes can be associated to lower quality
cells. Cells can also have lower expressed genes due to their biological state
or celltype.

```{r}
pct_cells<- 50
min_umi  <- 3000
```

Select clusters with `r pct_cells` % cells having less than `r min_umi`umi
counts.

```{r}
umi_df <- as.data.frame(cbind(sce$total, sce$cluster_k5))
colnames(umi_df) <- c("umi", "cluster_k5")

# caculate the percentatges of cells that are "low umi"
umi_df <- 
  umi_df %>% 
  mutate(low_umi = umi < min_umi) %>% 
  group_by(cluster_k5) %>% 
  summarise(n_low_umi = sum(low_umi), n_cells = n()) %>%
  mutate(pct_cell_low_umi = (n_low_umi/n_cells)*100) 
umi_df %>% arrange(desc(pct_cell_low_umi))

# Filter the clusters that have a high proportion of "low umi"
low_umi_clusters <- 
  umi_df %>% 
  filter(pct_cell_low_umi > pct_cells) %>% 
  .$cluster_k5
```

The clusters flagged are `r low_umi_clusters`

### Mithocondrial genes

High mithocondrial genes is associated with stressed, lower quality, cells.

```{r}
plotTSNE(sce, colour_by = "subsets_mt_percent") + 
  ggtitle("Percentatge mithocondrial genes")
```

```{r}
pct_cells<- 50
pct_mt  <- 10
```

Select clusters with `r pct_cells` % cells having more than `r pct_mt`%
mithocondrial genes.

```{r}
mt_df <- as.data.frame(cbind(sce$subsets_mt_percent, sce$cluster_k5))
colnames(mt_df) <- c("subsets_mt_percent", "cluster_k5")

mt_df <- 
  mt_df %>% 
  mutate(high_pct = subsets_mt_percent > pct_mt) %>% 
  group_by(cluster_k5) %>% 
  summarise(n_high_mt = sum(high_pct), n_cells = n()) %>% 
  mutate(pct_cell_high_mt = (n_high_mt/n_cells)*100)
mt_df %>% arrange(desc(pct_cell_high_mt))

high_mt_clusters <- 
  mt_df %>% 
  filter(pct_cell_high_mt > pct_cells) %>% 
  .$cluster_k5

```

The clusters flagged are `r high_mt_clusters`.


### Number of mice per cluster

How many mice contribute to each cluster?

```{r mice-per-cluster}
# count how many cells from each gnt group  there are per cluster
sum_per_gnt_cluster <- table(sce$genotype, sce$cluster_k5 )
# for each cluster count how many cells are from each mice, dividing by KO and WT mice
sum_per_mice_cluster <- table(sce$mouse, sce$genotype, sce$cluster_k5 )
# For each cluster sum of mice that do have cells on that cluster
colSums(sum_per_mice_cluster > 0)

# create a summary 
summary <- as.data.frame(rbind(colSums(sum_per_mice_cluster > 0), sum_per_gnt_cluster, colSums(sum_per_gnt_cluster)))
row.names(summary) <- c("KO mice", "WT mice", "KO cells", "WT cells", "total cells")
summary
```

Except from the obvious microglia clusters, where the numbers are very low or
even absent in the fire mice nothing stands out.

### Control vs fire mice

We want to have a closer look at the clusters that do have a difference between the knockout and the wild type before deleting the clusters.

```{r fig.width=10, message=FALSE}
# divide the two objects
sce_ko <- sce[,sce$genotype == "KO"]
sce_ctr <- sce[,sce$genotype == "WT"]
# plot them side by side
gridExtra::grid.arrange(
plotTSNE(sce_ko, colour_by = "cluster_k5", point_size=0.5,
         point_alpha = 0.3, text_by = "cluster_k5", text_size = 3) +
  scale_color_hue() +
  ggtitle("fire mice"),
plotTSNE(sce_ctr, colour_by = "cluster_k5", point_size=0.5,
         point_alpha = 0.3, text_by = "cluster_k5", text_size = 3) +
  scale_color_hue() +
  ggtitle("control"), 
ncol = 2
)
```

#### Proportion KO-WT

Visualise the proportions in a table

```{r}
prop_per_gnt <- round(prop.table(sum_per_gnt_cluster, margin = 2 )*100, 2)

prop_per_gnt<- as.data.frame(prop_per_gnt)
colnames(prop_per_gnt) <- c("Genotype", "cluster", "Proportion")

prop_per_gnt %>% 
  filter(Genotype == "KO") %>% 
  arrange(desc(Proportion))

```

visualise in a plot


```{r plot}
ggplot(data = prop_per_gnt, aes(x = cluster, y = Proportion, fill = Genotype)) +
  geom_bar(position = "fill", stat = "identity") + theme_classic() 

```

```{r}
ko_pct <- 60
```

```{r}
difference_KO_WT <-
  prop_per_gnt %>% 
  arrange(desc(Proportion)) %>% 
# the microglia clusters will be the top differences, but this is not 
    # something we are specially interested
  filter(!(cluster %in% c(21, 24, 1, 16))) %>% 
    filter(Genotype == "KO") %>% 
    # select the "interesting" clusters
  filter(Proportion > ko_pct ) %>% 
  .$cluster
```

The clusters with more than `r ko_pct` % cells from the KO animals are:

```{r}
gt_ko_pct <- unique(sce$cluster_k5[sce$cluster_k5 %in% c( high_mt_clusters, low_umi_clusters) &
    (sce$cluster_k5 %in% difference_KO_WT)])
```

`r gt_ko_pct`


 6 and 27 are very low quality non identified cells, most lickeley only clustering due to the quality and will be deleted. 
 
 37 is close to the big high mt. oligo cluster we are deleting. It does not have high mt but does have very low umi counts. This will also be deleted. 
 
32 and 23 are two small neuronal clusters (part of cluster 15 at k60) (approx. 70% KO). The other small neuronal clusters around, 36 and 12, would be deleted with the chosen thresholds (approx. 55% KO). However as cluster 15 (at k60) is the only neuronal cluster this whole cluster will be kept even though it is borderline in terms of quality. More information supporting this choice can be found in the cluster_QC_k60 and the first annotation files.

```{r}
# the cluster 15 from k60 is formed by the clusters bellow at k5 (cluster tree in clustering_01)
neuro <- c(12,32,23,35,36)
```

### Cluster QC

We filter out the clusters highlighted as:

-   low umi: majority of cells having less than 3000 umi counts

-   high mt : majority of cells having more than 10 % mitochondrial genes


```{r filterout}
sce$filter_out <- sce$cluster_k5 %in% c( high_mt_clusters, low_umi_clusters) &
                !(sce$cluster_k5 %in% neuro)
filter_out <- sum(sce$filter_out)
plotTSNE(sce, colour_by = "filter_out", point_size=0.5,
         point_alpha = 0.3, text_by = "cluster_k5", text_size = 3) 

if (!file.exists(here("processed", age, "sce_clusterQC.RDS"))){
 sce <- sce[sce$filter_out == FALSE, ]
 saveRDS(sce, here("processed", age, "sce_clusterQC.RDS"))
}


```

There are `r filter_out` cells filtered out.

## Session Info

<details>

<summary>

Click to expand

</summary>

```{r session-info}
sessionInfo()
```

</details>
