---
title: "Feature Selection and Dimensional reduction"
author: "NadineBestard"
date: "10/03/2021"
output: html_document
---

# Set-up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r set-up, message=FALSE, warning=FALSE}
library(SingleCellExperiment)
library(scran) # For feature selcetion
library(scater) # for pca
library(ggplot2) # modify plots
library(here) #reproducible paths
```

```{r}
age <- "old"
```

```{r load-sce}
sce <- readRDS(here("processed", age,  "sce_norm_02.RDS"))

```

# Feature selection

After the quality control the variable genes need to be selected again to better represent the variance in this cleaned dataset.
We follow the same methods than for our first feature selection, selecting here for 
the top 2000 genes.

```{r}
#The density weights are removed because the genes
# with highest mean abundance are also HVG, this avoids overfiting
gene_var_df <- modelGeneVar(sce, density.weights=FALSE )
gene_var <- metadata(gene_var_df)
plot(gene_var$mean, gene_var$var, xlab= "Mean of log-expression", ylab= "Variance of log-expression")
curve(gene_var$trend(x), lwd=2, add=T, col = "red")
# select hvgs
hvgs <- getTopHVGs(gene_var_df, n=2000)
# save them in the object
rowSubset(sce) <- hvgs
```

# Dimensionality reduction

## Run PCA and choose PCs
```{r}
set.seed(1000)
sce <- runPCA(sce)
```

```{r}
pct_var <- attr(reducedDim(sce, "PCA"), "percentVar")
plot(pct_var, log="y", xlab="PC", ylab="pct variance explained")
```


 will keep 26 PCs, as there is a drop between 25 and 26 PCs in 
the elbow plot. 

```{r}
#keep the previous dimensional reduction just in case
reducedDim(sce, "PCA_all") <- reducedDim(sce, "PCA")
# And replace the default PCA with the reduced PCs
reducedDim(sce, "PCA") <- reducedDim(sce, "PCA")[,1:26]
```

## Visualisation
For visualisation, reduce to 2 dimensions. Non linear reductions.

### UMAP
```{r}
sce <- runUMAP(sce, dimred="PCA")
plotReducedDim(sce, dimred="UMAP", colour_by="genotype", point_size=0.1, point_alpha = 0.3)
```

### tSNE
```{r}
sce <- runTSNE(sce, dimred="PCA")
plotReducedDim(sce, dimred="TSNE", colour_by="genotype", point_size=0.1, point_alpha = 0.3)
```

Visualise with the preliminary annotation.
```{r}
plotReducedDim(sce, dimred="TSNE", colour_by="celltype", text_by = "celltype", point_size=0.1, point_alpha = 0.3)
```

Visualise with the old clustering annotation. 
```{r}
plotReducedDim(sce, dimred="TSNE", colour_by="celltype", text_by = "cluster_k60", point_size=0.1, point_alpha = 0.3)
```

## Save objects
```{r}
saveRDS(sce, here("processed", age, "sce_dimred_02.RDS"))
```

## Session Info

<details>

<summary>Click to expand</summary>

```{r session-info}
sessionInfo()
```

</details>
